---
title: Uniform Manifold Approximation and Projection in R
output:
  html_document:
    toc: true
    toc_float: true
---

<style>
h1.title {
  margin-top: 1em;
  margin-bottom: 1.5em;
}
h2, h3 {
  margin-top: 2em;
  margin-bottom: 0.7em;
}
p {
  font-size: 18px;
  margin-bottom: 0.7em;
}
pre {
  font-size: 16px;
  line-height: 1.5;
}
</style>



```{r, echo=FALSE}
## block with some startup/background objects functions
library(umap)

iris.colors = c("#1f77b4", "#d62728", "#7f7f7f")
iris.colors = c("#ff7f00", "#e377c2", "#17becf")
plot.iris = function(x, labels,
         main="A UMAP visualization of the Iris dataset",
         pad=0.1, cex=0.65, pch=19, add=FALSE, legend.suffix="",
         cex.main=1, cex.legend=1) {

  layout = x
  if (class(x)=="umap") {
    layout = x$layout
  } 
  
  xylim = range(layout)
  xylim = xylim + ((xylim[2]-xylim[1])*pad)*c(-0.5, 0.5)
  if (!add) {
    par(mar=c(0.2,0.7,1.2,0.7), ps=10)
    plot(xylim, xylim, type="n", axes=F, frame=F)
    rect(xylim[1], xylim[1], xylim[2], xylim[2], border="#aaaaaa", lwd=0.25)  
  }
  points(layout[,1], layout[,2], col=iris.colors[as.integer(labels)],
         cex=cex, pch=pch)
  mtext(side=3, main, cex=cex.main)

  labels.u = unique(labels)
  legend.pos = "topright"
  legend.text = as.character(labels.u)
  if (add) {
    legend.pos = "bottomright"
    legend.text = paste(as.character(labels.u), legend.suffix)
  }
  legend(legend.pos, legend=legend.text,
         col=iris.colors[as.integer(labels.u)],
         bty="n", pch=pch, cex=cex.legend)
}

set.seed(123456)
```



## Introduction

Uniform Manifold Approximation and Projection (UMAP) is an algorithm for dimensional reduction proposed by [McInnes and Healy](https://arxiv.org/abs/1802.03426).

This vignette demonstrates how to use the `umap` R package to perform dimensional reduction and data trasnformations with the UMAP method. The vignette uses a small dataset as an example, but the package is suited to process larger data with many thousands of data points.




## Usage

For a practical demonstration, let's use the [Iris dataset](https://en.wikipedia.org/wiki/Iris_flower_data_set), `iris`.

```{r}
head(iris, 3)
```

The first four columns contain data and the last column contains a label. It will be useful to 
separate those components.

```{r}
iris.data = iris[, grep("Sepal|Petal", colnames(iris))]
iris.labels = iris[, "Species"]
```

### Creating a projection

Let's now load the `umap` package and apply the UMAP transformation.

```{r iris.umap}
library(umap)
iris.umap = umap(iris.data)
```

The output, `iris.umap`, is an object of class `umap`. We can get a minimal summary of its contents by just printing it.

```{r umap.print}
iris.umap
```

The main component of the object is `layout`, which holds a matrix with coordinates.

```{r umap.layout}
head(iris.umap$layout, 3)
```

These coordinates can be used to visualize the dataset. (The custom plot function, `plot.iris`, is available at the end of this vignette.)

```{r, fig.width=3.6, fig.height=3.6, dpi=150}
plot.iris(iris.umap, iris.labels)
```




### Projecting new data

Once we have a `umap` object describing a projection of one dataset into a low-dimensional layout, it becomes possible to project other data onto the same manifold. 

To perform the projection, we need a second dataset with the same data features as the training data. For this demonstration, let's create such data by adding some noise to the original data.

```{r}
iris.wnoise = iris.data + matrix(rnorm(150*40, 0, 0.1), ncol=4)
colnames(iris.wnoise) = colnames(iris.data)
head(iris.wnoise, 3)
```

We can now request to arrange these perturbed observations onto the same layout as before. Following R's design pattern for fitted models, this is performed via `predict`.
   
```{r}
iris.wnoise.umap = predict(iris.umap, iris.wnoise)
head(iris.wnoise.umap, 3)
```

The output here is a matrix with coordinates. We can now visualize this new data alongside the original.

```{r, fig.width=3.6, fig.height=3.6, dpi=150}
plot.iris(iris.umap, iris.labels)
plot.iris(iris.wnoise.umap, iris.labels, add=T, pch=4,
          legend.suffix=" (with noise)")
```

Note the new observations lie close to their original counterparts.




## Tuning UMAP

The example above uses function `umap` with a single argument - the input dataset - so the embedding is performed with default settings. However, the algorithm can be tuned in several ways. There are two strategies for tuning: via configuration objects and via additional arguments.


### Configuration objects

The default configuration object is called `umap.defaults`. This is a list encoding default values for all the parameters used within the algorithm.

```{r defaults, eval=FALSE}
umap.defaults
```

```{r defaults2, eval=TRUE, echo=FALSE, collapse=TRUE}
umap.defaults
```	

This object is a list with key-value pairs shown. To obtain some minimal information about each field, see the documentation in `help(umap.defaults)`, or see the original publication.


To create a custom configuration, create a copy of the defaults and then update any of the fields. For example, let's change the seed for random number generation.

```{r custom.config, eval=TRUE}
custom.config = umap.defaults
custom.config$random_state = 123
```

We can observe the changed settings by inspecting the object again (try it). To perform the UMAP projection with these settings, we can run the projection again and pass the configuration object as a second argument.

```{r custom2, fig.width=3.6, fig.height=3.6, dpi=150}
iris.umap.2 = umap(iris.data, custom.config)
plot.iris(iris.umap.2, iris.labels,
          main="Another UMAP visualization (different seed)")
```

The result is slightly different than before due to a new instantiation of the random number generator.



### Additional arguments

Another way to customize the algorithm is to specify the non-default parameters explicitly. To achieve equivalent results to the above, we can thus use

```{r custom3, eval=FALSE}
iris.umap.3 = umap(iris.data, random_state=123)
```

The coordinates in this new output object should match the ones from `iris.umap.2` (check it!)




## Implementations

The package provides two implementations of the umap method, one written in R and one accessed via an external python module.

The implementation written in R is the default. This implementation follows the design principles of the UMAP algorithm and its running time scales better-than-quadratically with the number of items (points) in a dataset. It is thus in principle suitable for use on datasets with thousands of points. It is the default because it should be functional without extensive dependencies.

The second available implementation is a wrapper for a python package `umap-learn`. To enable this implementation, specify the argument `method`,

```{r, eval=FALSE}
iris.umap.4 = umap(iris.data, method="umap-learn")
```

This command has several dependencies. To make it work, you must have the `reticulate` package installed and loaded (use `install.packages("reticulate")` and `library(reticulate)`). Furthermore, you must have the `umap-learn` python package installed (see the [package repo](https://github.com/lmcinnes/umap) for instructions). If either of these components is not available, the above command will display an error message.

Note that it will not be possible to produce exactly the same output from the two implementations due to inequivalent random number generators in R and python, and due to slight discrepancies in the implementations.


&nbsp;

## Appendix

The custom plot function used to visualize the Iris dataset:

```{r show.plot.iris}
plot.iris
```



Summary of R session:

```{r}
sessionInfo()
```

&nbsp;