---
title: Uniform Manifold Approximation and Projection in R
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{UMAP}
  %\usepackage[UTF-8]{inputenc}
---

<style>
h1.title {
  margin-top: 1em;
  margin-bottom: 1.5em;
}
h2, h3 {
  margin-top: 2em;
  margin-bottom: 0.7em;
}
p {
  font-size: 18px;
  margin-bottom: 0.7em;
}
pre {
  font-size: 16px;
  line-height: 1.5;
}
</style>



```{r, echo=FALSE}
## block with some startup/background objects functions

iris.colors = c("#1f77b4", "#d62728", "#7f7f7f")
iris.colors = c("#ff7f00", "#e377c2", "#17becf")
plot.iris = function(x, labels,
         main="A UMAP visualization of the Iris dataset",
         pad=0.02, cex=0.65, pch=19,
         cex.main=1, cex.legend=1) {

  layout = x$layout
  par(mar=c(0.2,0.7,1.2,0.7), ps=10)
  xylim = range(layout)
  xylim = xylim + ((xylim[2]-xylim[1])*pad)*c(-0.5, 0.5)
  plot(xylim, xylim, type="n", axes=F, frame=F)
  xylim = par()$usr
  rect(xylim[1], xylim[1], xylim[2], xylim[2], border="#aaaaaa", lwd=0.2)
  points(layout[,1], layout[,2], col=iris.colors[as.integer(labels)],
         cex=cex, pch=pch)
  mtext(side=3, main, cex=cex.main)

  labels.u = unique(labels)
  legend("topright", legend=as.character(labels.u),
         col=iris.colors[as.integer(labels.u)],
         bty="n", pch=pch, cex=cex.legend)
}

set.seed(123456)
```



## Introduction

Uniform Manifold Approximation and Projection (UMAP) is an algorithm for dimensional reduction proposed by [McInnes and Healy](https://arxiv.org/abs/1802.03426). This vignette demonstrates how to use the `umap` R package to perform dimensional reduction and visualization with the UMAP method.


## Usage

For a practical demonstration, let's use the Iris dataset. This dataset is accessible through object `iris`.

```{r}
head(iris, 3)
```

The first four columns contain data, the last column contains a label. It will be useful to 
separate those components.

```{r}
iris.data = iris[, grep("Sepal|Petal", colnames(iris))]
iris.labels = iris[, "Species"]
```

Now, let's load the `umap` package and apply the UMAP transformation.

```{r iris.umap, cache=TRUE}
library(umap)
iris.umap = umap(iris.data)
```

The output is here an object; printing it gives a minimal summary of its contents.

```{r umap.print}
iris.umap
```

The main compoent of the object is `layout`, which holds a matrix with coordinates.

```{r umap.layout}
head(iris.umap$layout, 3)
```

We can now use these coordinates to visualize the dataset on a 2D plane. 

```{r, fig.width=3.6, fig.height=3.6, dpi=150}
plot.iris(iris.umap, iris.labels)
```

(The custom plot function, `plot.iris`, is available at the end of this vignette.)




## Tuning UMAP

The example above uses function `umap` with a single argument - the input dataset - so the embedding is performed with default settings. However, the algorithm can be tuned in several ways. There are two strategies for tuning: via configuration objects and via additional arguments.


### Configuration objects

The default configuration object is called `umap.defaults`. This is a list encoding default values for all the parameters used within the algorithm.

```{r defaults, eval=FALSE}
umap.defaults
```

```{r defaults2, eval=TRUE, echo=FALSE, collapse=TRUE}
umap.defaults
```	

This object is a list with key-value pairs shown. To obtain some minimal information about each field, see the documentation in `help(umap.defaults)`, or see the original publication.


To create a custom configuration, create a copy of the defaults and then update any of the fields. For example, let's change the seed for random number generation.

```{r custom.config, eval=TRUE}
custom.config = umap.defaults
custom.config$seed = 123
```

We can observe the changed settings by inspecting the object again (try it). To perform the UMAP projection with these settings, we can run the projection again and pass the configuration object as a second argument.

```{r custom2, cache=TRUE, fig.width=3.6, fig.height=3.6, dpi=150}
iris.umap.2 = umap(iris.data, custom.config)
plot.iris(iris.umap.2, iris.labels,
          main="Another UMAP visualization of the Iris dataset (different seed)")
```

The result is slightly different due to a new instanciation of the random number generator.



### Additional arguments

Another way to customize the algorithm is to specify the non-default parameters explicitly. To achieve equivalent results to the above, we can thus use

```{r custom3, eval=FALSE}
iris.umap.3 = umap(iris.data, seed=123)
```

The coordinates in this new output object should match the ones from `iris.umap.2` (check it!)




## Performance

The implementation in the `umap` package follows the design principles of the UMAP algorithm. As such, the embedding transformation scales better-than-quadratically with the number of items (points) in a dataset. It is thus in principle suitable for use on datasets with thousands of points.

The current implementation is written in pure R with only occasional use of Rcpp. Some parts are not optimized for speed. Therefore, absolute performance may lag behind implementations using precompiled code.




&nbsp;

## Appendix

The custom plot function used to visualize the Iris dataset:

```{r show.plot.iris}
plot.iris
```



Summary of R session:

```{r}
sessionInfo()
```

&nbsp;