---
title: Uniform Manifold Approximation and Projection in R
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{UMAP}
  %\usepackage[UTF-8]{inputenc}
---

<style>
h1.title {
  margin-top: 1em;
  margin-bottom: 1.5em;
}
h2, h3 {
  margin-top: 2em;
  margin-bottom: 0.7em;
}
p {
  font-size: 18px;
  margin-bottom: 0.7em;
}
pre {
  font-size: 16px;
  line-height: 1.5;
}
</style>



```{r, echo=FALSE}
## block with some startup/background objects functions

iris.colors = c("#1f77b4", "#d62728", "#7f7f7f")
iris.colors = c("#ff7f00", "#e377c2", "#17becf")
plot.iris = function(layout, labels,
                     main="A UMAP visualization of the Iris dataset",
                     pad=0.02, cex=0.65, pch=19,
                     cex.main=1, cex.legend=1) {
  par(mar=c(0.2,0.7,1.2,0.7), ps=10)
  xylim = range(layout)
  xylim = xylim + ((xylim[2]-xylim[1])*pad)*c(-0.5, 0.5)
  plot(xylim, xylim, type="n", axes=F, frame=F)
  xylim = par()$usr
  rect(xylim[1], xylim[1], xylim[2], xylim[2], border="#aaaaaa", lwd=0.2)
  points(layout[,1], layout[,2], col=iris.colors[as.integer(labels)],
         cex=cex, pch=pch)
  mtext(side=3, main, cex=cex.main)

  labels.u = unique(labels)
  legend("topright", legend=as.character(labels.u),
         col=iris.colors[as.integer(labels.u)],
          bty="n", pch=pch, cex=cex.legend)
  
}

set.seed(123456789)
```



## Introduction

UMAP stands for Uniform Manifold Approximation and Projection. It is an algorithm for dimensional reduction proposed by [McInnes and Healy](https://arxiv.org/abs/1802.03426).

[To do.]


This vignette demonstrates how to use the `umap` R package to perform dimensional reduction
and visualization.


## Usage

To begin, let's load the package.

```{r}
library(umap)
```

For a practical demonstration, let's use the Iris dataset. This dataset is accessible through object `iris`.

```{r}
head(iris, 3)
```

The first four columns contain data, the last column contains a label. It will be useful to 
separate the data from the labels, so let's create objects for those components.

```{r}
iris.data = iris[, grep("Sepal|Petal", colnames(iris))]
iris.labels = iris[, "Species"]
```

Let's now apply the UMAP transformation.

```{r iris.umap, cache=TRUE}
iris.umap = umap(iris.data)
```

The output is here a matrix with coordinates for each point in the original dataset.
We can now visualize the dataset on a 2D plane. 

```{r, fig.width=4, fig.height=4, dpi=150}
plot.iris(iris.umap, iris.labels)
```

(The custom plot function, `plot.iris`, is available at the end of this vignette.)




## Tuning UMAP

The example above uses function `umap` with a single argument - the input dataset. The transformation is performed with default settings. However, the algorithm can be tuned in several ways. There are two strategies for tuning: via configuration objects and via additional arguments.


### Configuration objects

The default configuration object is called `umap.defaults`. This is a list encoding default values for all the parameters used within the algorithm.

```{r defaults, eval=FALSE}
umap.defaults
```

```{r defaults2, eval=TRUE, echo=FALSE, collapse=TRUE}
umap.defaults
```	

This object is a list with key-value pairs shown. To obtain some minimal information about each of field, see the documentation in `help(umap.defaults)`, or see the original publication.


To create a custom configuration, create a copy of the defaults and then update any of the fields. For example, let's change the seed for random number generation.

```{r custom.config, eval=TRUE}
custom.config = umap.defaults
custom.config$seed = 123123
```

We can observe the changed settings by inspecting the object again

```{r, eval=FALSE}
custom.config
```
```{r, eval=TRUE, echo=FALSE, collapse=TRUE}
custom.config
```

To perform the UMAP projection with these settings, we can run the projection again and pass the configuration object as a second argument.

```{r custom2, fig.width=4, fig.height=4, dpi=150}
iris.umap.2 = umap(iris.data, custom.config)
plot.iris(iris.umap.2, iris.labels,
          main="Another UMAP visualization (different seed)")
```




### Additional arguments

Another way to customize the algorithm is to specify the non-default parameters explicitly. To achieve equivalent results to the above, we can thus use

```{r custom3, eval=FALSE}
iris.umap.3 = umap(iris.data, seed=123123)
```

The coordinates in this new output object should match the ones from `iris.umap.2` (check it!)






&nbsp;

## Appendix

The custom plot function used to visualize the Iris dataset:

```{r show.plot.iris}
plot.iris
```



Summary of R session:

```{r}
sessionInfo()
```

&nbsp;